Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2008-11-26 docs/pubs/ISEI6/main.tex

\newcommand{\owlifier}{Owlifier}
\newcommand{\mypara}[1]{\vspace{12pt}\noindent\textbf{#1}}

@@

\begin{abstract}
abstract here...
\end{abstract}

@@

The use of formal ontologies can significantly enhance metadata
descriptions of ecological data by providing richer and more
consistent terminologies. Annotating data with ontology terms can both
help users interpret data as well as enable advanced capabilities for
data discovery and integration, e.g., by exploiting subsumption and
part-of hierarchies as well as more formal constraints such as
cardinality restrictions on properties and term equivalence. Many
efforts are underway to engage scientists in the development of
ontologies to represent common ecological and biodiversity
concepts. Most of these projects leverage OWL (the W3C Web Ontology
Language) as a standard XML syntax for representing and sharing
ontologies. A key advantage of OWL is that it is supported by a wide
range of generic tools, including editors, reasoning systems, query
languages, and storage technologies. These tools, however, are
primarily targeted at experts in knowledge engineering and software
development. This is especially true with ontology editors (such as
Prot?g?, OWLed, SWOOP, etc.), which allow for very detailed ontology
specifications, but at the same time require a considerable amount of
understanding of the underlying ontology formalisms and syntax. Thus,
we see the lack of suitable ontology editing tools for scientists as
one of the major barriers for more wide-scale adoption of ontologies
in ecology.

Here, we present a novel approach for ontology creation that aims at
being more intuitive for ecologists and that can be used to rapidly
construct large ontologies for describing scientific data. Our
approach is to allow scientists to use common spreadsheet tools to
describe, in an intuitive way, different aspects of an ontology, and
then to take these descriptions and convert them into full-fledged OWL
ontologies using a software application called owlifier. An owlifier
spreadsheet consists of a set of "blocks" that have a predefined
template structure for users to fill in. Each non-empty row in an
owlifier table constitutes a block, where blocks can be used to define
ontology concepts (including sub-concepts), concept synonyms,
overlapping concepts (i.e., that by definition are not disjoint),
properties (including their domain and range), and inverse
properties. We also provide blocks for plain-text descriptions of
concepts and properties, and for referencing one or more existing
ontologies (e.g., to extend an ontology or to define ontology
articulations). Blocks can be sparse (inheriting from previous
blocks), which simplifies the creation of large ontologies. Although
not as expressive as full-fledged OWL ontologies, our approach can
produce ontology structures essential for enhanced data discovery and
integration, while at the same time providing a more accessible user
interface for ecologists. Further, compared to many existing editors,
our approach can be used to rapidly construct concept hierarchies,
e.g., from long lists of keywords, within familiar spreadsheet
software. We have field-tested the owlifier approach with ecologists
and evolutionary biologists working with trait data, and found this
application enabled them to quickly and easily comprehend and
construct ontologies.

@@

\mypara{Import Block}

\mypara{}Import blocks assign namespace labels to external
ontologies. Each external ontology is imported into the current
ontology. We refer to the ontologies of import blocks as
\emph{imported ontologies}.  Using import blocks, concepts and
properties of imported ontologies can be used within other blocks of
the table.  Import blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline
    \texttt{import} & $\textit{N}$ & $\textit{U}$ \\ \hline 
  \end{tabular} 
\end{itemize}
where $\textit{N}$ is a namespace label and $\textit{U}$ is an OWL
ontology URI. Classes and properties from imported ontologies are
referenced by prefixing the namespace label $\textit{N}$ to the
corresponding class or property name in the normal way. The following
block imports the ``earth realm'' ontology from the set of SWEET
ontologies \cite{raskin:_seman_web_for_earth_and}.
\begin{tabbing}
  ~~\texttt{import sweet
    http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
\end{tabbing}
where the ontology class denoting marine ecosystems can be refered to
from within the corresponding \owlifier\ table using the syntax
\texttt{sweet:MarineEcosystem}. This class is referred to an an
``imported'' class.

\mypara{Class Block}

\mypara{}Class blocks specify class and subclass relationships. Class
blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{class} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where each class $C_i$ is asserted in the current ontology to subsume
a class $C_{i+1}$, for $1 \le i < n$. Each $C_i$ in an entity block
induces a DL axiom \[C_i \sqsubseteq C_{i+1}.\] If both $C_i$ and
$C_{i+1}$ are imported classes, we say that the block defines an
``articulation'' (i.e., mapping) between them. The following class
block defines a simple subclass hierarchy
\begin{tabbing}
~~\texttt{class PhysicalFeature AquaticPhysicalFeature River}
\end{tabbing}
stating that a river is a subclass of aquatic physical features and
that aquatic physical features are a type of physical feature.

\mypara{Synonym Block}

\mypara{}Synonym blocks define an equivalence relationship between
classes.  A synonym block has the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{synonym} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each class $C_i$ is asserted as being equivalent to class
$C_{i+1}$ in the current ontology, for $1 \le i < n$. For example, the
block
\begin{tabbing}
~~\texttt{synonym Maize Corn}
\end{tabbing}
defines maize and corn as synonym (i.e., equivalent) classes.

>>>>

\newcommand{\owlifier}{\textsf{Owlifier}}
\newcommand{\mypara}[1]{\vspace{12pt}\noindent\textbf{#1}}

@@

\begin{abstract}
  Formal ontlogies promise to provide many added advantages to
  ecological data management ... Most ontology editing frameworks and
  tools, however, are design for knowledge-representation experts and
  those with considerable understanding of the complexities of
  ontology languages (i.e., description logics). This paper proposes
  an approach for ontology creation based on ...
\end{abstract}

@@

The use of formal ontologies can significantly enhance metadata
descriptions of ecological data by providing richer and more
consistent terminologies \cite{tree,arees,parr}. Annotating data with
ontology terms can both help users interpret data as well as enable
advanced capabilities for data discovery and integration, e.g., by
exploiting subsumption and part-of hierarchies as well as more formal
constraints such as cardinality restrictions on properties and term
equivalence \cite{ecoinf,other?}. Many efforts are underway to engage
scientists in the development of ontologies to represent common
ecological and biodiversity concepts. Most of these projects leverage
the W3C Web Ontology Language (OWL) as a standard XML syntax for
representing and sharing ontologies. A key advantage of OWL is that it
is supported by a wide range of generic tools, including editors
\cite{protege,swoop}, reasoning systems \cite{pellet,fact++}, query
languages \cite{???}, and storage technologies \cite{jena,???}. Most
of these tools, however, are primarily targeted at experts in
knowledge engineering and software development. This is especially
true with ontology editors (such as Protege, OWLed, SWOOP, etc.),
which allow for very detailed ontology specifications, but at the same
time require a considerable amount of understanding of the underlying
ontology formalisms and syntax. Thus, we see the lack of suitable
ontology editing tools for scientists as one of the major barriers for
more wide-scale adoption of ontologies in ecology.

Here, we present a novel approach for ontology creation that aims at
being more intuitive for ecologists and that can be used to rapidly
construct large ontologies for describing scientific data. Our
approach is to allow scientists to use common spreadsheet-based tools
to describe, in an intuitive way, different aspects of an ontology,
and then to take these descriptions and convert them into full-fledged
OWL ontologies using a software application called \owlifier.

An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have
a predefined template structure for users to fill in. Each non-empty
row in an owlifier table constitutes a block, where blocks can be used
to define ontology classes (including subclasses), class synonyms,
data- and object-type properties (and constraints), etc.  We also
provide blocks for plain-text descriptions of concepts and properties,
and for referencing one or more existing ontologies (e.g., to extend
an existing ontology or to define ontology articulations). Blocks can
be sparse (inheriting from previous blocks), which simplifies the
creation of large ontologies.

Although not as expressive as full-fledged OWL ontologies, our
approach can produce ontology structures essential for enhanced data
discovery and integration, while at the same time providing a more
accessible user interface for ecologists. Further, compared to many
existing editors, our approach can be used to rapidly construct
concept hierarchies, e.g., from long lists of keywords, within
familiar spreadsheet software. We have field-tested the owlifier
approach with ecologists and evolutionary biologists working with
trait data, and found this application enabled them to quickly and
easily comprehend and construct ontologies.

In this paper, we describe the syntax and semantics of \owlifier. We
consider a set of blocks that generally follow the ...

@@

\mypara{Import Block.} Import blocks assign namespace labels to
external ontologies. Each external ontology is imported into the
current ontology. We refer to the ontologies of import blocks as
\emph{imported ontologies}.  Using import blocks, classes and
properties of imported ontologies can be used within other blocks of
an \owlifier\ table.  Import blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline
    \texttt{import} & $\textit{N}$ & $\textit{U}$ \\ \hline 
  \end{tabular} 
\end{itemize}
where $\textit{N}$ is a namespace label and $\textit{U}$ is an OWL
ontology URI. Classes and properties from imported ontologies are
referenced by prefixing the namespace label $\textit{N}$ to the
corresponding class or property name in the normal way. The following
block imports the ``earth realm'' ontology from the set of SWEET
ontologies \cite{raskin:_seman_web_for_earth_and}
\begin{tabbing}
  ~~\texttt{import sweet
    http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
\end{tabbing}
where the ontology class denoting marine ecosystems can be refered to
from within the corresponding \owlifier\ table using the syntax, e.g.,
\sloppy\texttt{sweet:MarineEcosystem}. Because this class refers to a
class in another ontology, we call it an ``imported'' class.

\mypara{Concept Block.} Concept blocks introduce new OWL classes and
specify subclass relationships. Imported classes may also be used
within a concept block by prefixing the class name with a namespace
label, e.g., when an ontology is being extended.  We use the term
`concept' to avoid confusion with the use of the term 'class' in
biological taxonomies and because `concept' is often more intuitive
for users. Concept blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{concept} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where each class $C_i$ is asserted in the current ontology to subsume
a class $C_{i+1}$, for $1 \le i < n$. Each $C_i$ in a concept block
induces the DL axiom \[C_i \sqsubseteq C_{i+1}.\] If both $C_i$ and
$C_{i+1}$ are imported classes, we say that the block defines an
``articulation'' (i.e., mapping) between the two classes. The
following concept block defines a simple subclass hierarchy
\begin{tabbing}
  ~~\texttt{concept PhysicalFeature AquaticPhysicalFeature River}
\end{tabbing}
which states that river is a subclass of aquatic physical feature and
that aquatic physical feature is a subclass of physical feature.

\mypara{Synonym Block.} Synonym blocks define an equivalence
relationship between classes.  A synonym block has the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{synonym} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each class $C_i$ is asserted as being equivalent to class
$C_{i+1}$ in the current ontology, for $1 \le i < n$. For example, the
block
\begin{tabbing}
~~\texttt{synonym Maize Corn}
\end{tabbing}
defines maize and corn as synonym (i.e., equivalent) classes.
